# Nonparametric Estimation

## Slides {.unnumbered}

Chapter slides [here](chap3.html){target="_blank"}. (To convert html to pdf, press E $\to$ Print $\to$ Destination: Save to pdf)

## R-code {.unnumbered}

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false

##################################################################
# This code generates the numerical results in chapter 3         #
##################################################################

# install.packages("rmt") # if not already installed
library(rmt)
library(tidyverse)

##### Read in HF-ACTION DATA ########
data(hfaction)
head(hfaction)
#> Displays the first few rows of the HF-ACTION dataset (patid, time, status, trt_ab)

# TFE: take the first event per patient id
# TFE = Time to First Event (death or hospitalization)
hfaction_TFE <- hfaction |>
  arrange(patid, time) |>
  group_by(patid) |>
  slice_head() |>
  ungroup()

###### Standard RMST analysis #################
library(survRM2)

# -------------------------------------------------------------------------
# Mortality analysis
# -------------------------------------------------------------------------

## get mortality data
hfaction_D <- hfaction |>
  filter(status != 1) # remove hospitalization records (status=1)

## RMST (Restricted Mean Survival Time) analysis for overall survival
# Here we compare arm=1 (training) vs arm=0 (usual care)
rmst_obj <- rmst2(
  time    = hfaction_D$time,
  status  = hfaction_D$status > 0,
  arm     = hfaction_D$trt_ab,
  tau     = 3.97
)
rmst_obj
#> Prints estimates of restricted mean survival (in years), differences, ratios, and p-values
# Between-group contrast 
#                       Est. lower .95 upper .95     p
# RMST (arm=1)-(arm=0) 0.238     0.013     0.464 0.039
# RMST (arm=1)/(arm=0) 1.074     1.003     1.150 0.040
# RMTL (arm=1)/(arm=0) 0.680     0.468     0.988 0.043

# Extract the difference in RMST (in years) and convert to months
rmst <- rmst_obj$unadjusted.result[1, 1] * 12 
rmst_p <- rmst_obj$unadjusted.result[1, 4]

# -------------------------------------------------------------------------
# TFE analysis
# -------------------------------------------------------------------------

## Check how many of the first events are death (status=1) or hospitalization (status=2)
hfaction_TFE |>
  count(status)

## RMST analysis for hospitalization-free survival
rmest_obj <- rmst2(
  time    = hfaction_TFE$time,
  status  = hfaction_TFE$status > 0,
  arm     = hfaction_TFE$trt_ab,
  tau     = 3.97
)
rmest_obj
#> Similar output for restricted mean event-free survival
# Between-group contrast 
#                       Est. lower .95 upper .95     p
# RMST (arm=1)-(arm=0) 0.198    -0.064     0.459 0.139
# RMST (arm=1)/(arm=0) 1.145     0.957     1.370 0.139
# RMTL (arm=1)/(arm=0) 0.924     0.832     1.027 0.141

# Extract the difference in event-free survival (years) and convert to months
rmest <- rmest_obj$unadjusted.result[1, 1] * 12
rmest_p <- rmest_obj$unadjusted.result[1, 4]

# -------------------------------------------------------------------------
# Mortality vs TFE (Kaplan-Meier plots with annotations)
# -------------------------------------------------------------------------

library(ggsurvfit)
library(patchwork)

# Plot for overall survival
pD <- survfit2(Surv(time, status > 0) ~ trt_ab, data = hfaction_D) |>
  ggsurvfit(linewidth = 1) +
  scale_ggsurvfit() +
  annotate(
    "text", x = 4, y = 1, hjust = 1, vjust = 1,
    label = str_c("4y-RMST = ", round(rmst, 2), " months",
                  " (P = ", round(rmst_p, 3), ")")
  ) +
  scale_color_discrete(labels = c("Usual care", "Training")) +
  scale_x_continuous("Time (years)", limits = c(0, 4)) +
  labs(y = "Overall survival")

# Plot for hospitalization-free survival
pTFE <- survfit2(Surv(time, status > 0) ~ trt_ab, data = hfaction_TFE) |>
  ggsurvfit(linewidth = 1) +
  annotate(
    "text", x = 4, y = 1, hjust = 1, vjust = 1,
    label = str_c("4y-RMEST = ", round(rmest, 2), " months",
                  " (P = ", round(rmest_p, 3), ")")
  ) +
  scale_ggsurvfit() +
  scale_color_discrete(labels = c("Usual care", "Training")) +
  scale_x_continuous("Time (years)", limits = c(0, 4)) +
  labs(
    y = "Hospitalization-free survival",
    caption = "RMEST: restricted mean event-free survival time"
  )

# Combine side-by-side
pD + pTFE + plot_layout(guides = "collect") &
  theme(
    legend.position = "top",
    legend.text = element_text(size = 12)
  ) 

# # ggsave("images/est_hfaction_unis.png", width = 8, height = 4.6)
#> Uncomment above to save the combined figure

# -------------------------------------------------------------------------
# RMT-IF analysis
# -------------------------------------------------------------------------

# rmtfit() from the rmt package fits the restricted mean time lost / 
# time free approach for recurrent events & death
obj <- rmtfit(rec(patid, time, status) ~ trt_ab, data = hfaction)

summary(obj, Kmax = 1, tau = 3.97)
#> Summarizes the model up to a maximum follow-up of 3.97 years, 
#> focusing on the first event (Kmax=1) or aggregated events

#############################################################
# Graphical analysis of the HF-ACTION trial to
# evaluate the effect of exercise training.
###########################################################

par(mfrow = c(1, 2))

# bouquet() displays the "bouquet plot" for each k up to Kmax=4
bouquet(
  obj, Kmax = 4, cex.group = 1.0,
  xlab = "Restricted mean win/loss time (years)",
  ylab = "Follow-up time (years)",
  group.label = FALSE, ylim = c(0, 4.2)
)
text(-0.8, 4.15, paste("Usual care"))
text(0.8, 4.15, paste("Exercise training"))

# plot() visualizes the RMT-IF difference as a function of time
plot(
  obj, conf = TRUE, lwd = 2,
  xlab = "Follow-up time (years)",
  ylab = "RMT-IF of training (years)", 
  main = ""
)

par(mfrow = c(1, 1))
#> Reset the graphic device to a single panel

### LaTeX table ###

pval_fmt3 = function(x) {
  if (x < 0.001) {
    return("$<$0.001")
  } else {
    return(round(x, 3))
  }
}

ltable = NULL
# aggregate the results for k=1,...,K
hosp_sum = summary(obj, Kmax = 1, tau = 3.97)$tab
# aggregate the results for k=4,...,K
all_sum = summary(obj, Kmax = 4, tau = 3.97)$tab

ltable = c(
  "&", "&", "&",
  round(12 * hosp_sum[1, 1], 2), "&", round(12 * hosp_sum[1, 2], 2),
  "&", pval_fmt3(hosp_sum[1, 4]), "\\"
)

for (i in 1:6) {
  tmp = c(
    "&", i, "&&",
    round(12 * all_sum[i, 1], 2), "&", round(12 * all_sum[i, 2], 2),
    "&", pval_fmt3(all_sum[i, 4]), "\\"
  )
  ltable = rbind(ltable, tmp)
}

ltable[5, 2] = "4+"
ltable[6:7, 2] = ""

rownames(ltable) = c("Hopitalization", "", "", "", "", "Death", "Overall")

noquote(ltable)
#> Produces a LaTeX-formatted table summarizing the RMT-IF results

#######################################################################
#               WA analysis                                           #
#######################################################################

# install.packages("WA") # if needed
library(WA)

# load the hf-action study data (already loaded, but repeated here)
head(hfaction)

# descriptive analysis

## death & hosp rates by treatment arm
hfaction |>
  group_by(trt_ab, patid) |>
  summarize(
    ND = sum(status == 2),
    NH = sum(status == 1)
  ) |>
  summarize(
    death_rate = mean(ND),
    avgNH = mean(NH),
    sdNH = sd(NH)
  )

# weighted while-alive event rate analysis 
# with death weighted as 2 vs 1 for hospitalization
obj <- LRfit(
  id = hfaction$patid,
  time = hfaction$time,
  status = hfaction$status,
  trt = hfaction$trt_ab,
  Dweight = 2
)
## print some descriptive information 
obj

## summarize the inference results at tau=4 years
summary(obj, tau = 3.97, joint.test = TRUE)

plot(obj)
#> Plots the Weighted-Alive event rates over time for each arm

# -------------------------------------------------------------------------
# unadjusted cumulative mean (Ch 1)
# -------------------------------------------------------------------------

## fit proportional means model with death = 2 x hosp
library(Wcompo)

## change status coding
status <- hfaction$status
status[status != 0] <- 3 - status[status != 0]
#> This transforms status=1->2, status=2->1, effectively swapping them

obj_ML <- CompoML(
  hfaction$patid,
  hfaction$time,
  status,
  hfaction$trt_ab,
  w = c(2, 1)
)
## summary results
t <- obj_ML$t     # time points
mu0 <- obj_ML$y   # baseline cumulative mean events
mu1 <- mu0 * exp(as.numeric(obj_ML$beta))
#> The exponentiated beta shifts the baseline function for the second arm

## plot survival-adjusted cumulative event 
## vs unadjusted under PM model
plot(
  obj,
  ylab = "Cumulative loss",
  xlab = "Time (years)"
)
lines(
  t[t <= 3.97], mu0[t <= 3.97],
  lty = 3, col = "red", lwd = 2
)
lines(
  t[t <= 3.97], mu1[t <= 3.97],
  lty = 3, col = "blue", lwd = 2
)
legend(
  "bottomright",
  col = c("red", "red", "blue", "blue"),
  c("Usual care (WA)", "Usual care (unadj)",
    "Training (WA)", "Training (unadj)"),
  lty = c(1, 3, 1, 3),
  lwd = 2
)
#> Compares Weighted-Alive-based event accumulation with the "unadjusted" PM model


```

$$\newcommand{\d}{{\rm d}}
\newcommand{\T}{{\rm T}}
\newcommand{\dd}{{\rm d}}
\newcommand{\cc}{{\rm c}}
\newcommand{\pr}{{\rm pr}}
\newcommand{\var}{{\rm var}}
\newcommand{\se}{{\rm se}}
\newcommand{\indep}{\perp \!\!\! \perp}
\newcommand{\Pn}{n^{-1}\sum_{i=1}^n}
\newcommand\mymathop[1]{\mathop{\operatorname{#1}}}
\newcommand{\Ut}{{n \choose 2}^{-1}\sum_{i<j}\sum}
\def\a{{(a)}} 
\def\b{{(1-a)}} 
\def\t{{(1)}} 
\def\c{{(0)}} 
\def\d{{\rm d}} 
\def\T{{\rm T}} 
\def\bs{\boldsymbol}
$$

## Restricted Win Ratio

The win ratio estimand depends on the censoring distribution, as each pairwise comparison is limited to the observed follow-up $[0, C_i^\t \wedge C_j^\c]$. This leads to a censoring-weighted average of time-specific win probabilities, which is trial-dependent and lacks generalizability. One solution is to pre-specify a restriction time $\tau$ so all comparisons are evaluated over the same window.

For a two-tiered composite of death and a nonfatal event, the restricted win/loss probability is defined by \begin{align}
w_{a, 1-a}(\tau) &= \pr\{D^\b < \min(D^\a, \tau)\} \\
&\quad + \pr\{\min(D^\t, D^\c) > \tau, T^\b < \min(T^\a, \tau)\},
\end{align} with the restricted win ratio defined as $\text{WR}(\tau) = w_{1,0}(\tau) / w_{0,1}(\tau)$. Estimation methods include inverse probability of censoring weighting (IPCW) and multiple imputation (MI), as implemented in the `WINS` package.

## Restricted Mean Time in Favor (RMT-IF)

The restricted mean time in favor (RMT-IF) redefines the win-loss comparison in terms of time spent in more favorable states. For a multistate outcome $Y(t)$, RMT-IF measures the net average time one subject remains in a better state than another: $$
\mu(\tau) = E\left[ \int_0^\tau I\{Y^{(1)}(t) < Y^{(0)}(t)\} \, \dd t \right] 
- E\left[ \int_0^\tau I\{Y^{(0)}(t) < Y^{(1)}(t)\} \, \dd t \right].
$$

The process $Y(t)$ is assumed progressive and represents a hierarchy of worsening states, terminating in death. The estimand $\mu(\tau)$ admits a decomposition into component-specific effects: $$
\mu(\tau) = \sum_{k=1}^K \mu_k(\tau) + \mu_\infty(\tau),
$$ where each $\mu_k(\tau)$ captures net time gained in a specific nonfatal state and $\mu_\infty(\tau)$ captures the difference in restricted mean survival time. Estimation proceeds by plugging in Kaplan–Meier estimators for transition times. The `rmt::rmtfit()` function provides estimation, inference, and graphics.

## While-Alive Weighted Loss

Standard analyses of weighted total events may misrepresent treatment effects if survival time differs between groups. The while-alive loss framework defines a normalized event rate over time alive: $$
\ell^{(a)}(\tau) = \frac{E\left\{N_{\rm R}^{*(a)}(\tau)\right\}}{E\left(D^{(a)} \wedge \tau\right)}.
$$

More generally, the loss can be specified by a function $\mathcal{L}(\mathcal{H}^*(t))$ that accumulates only during time alive. This leads to a class of estimands: $$
\ell^{(a)}(\tau) = \frac{E\left\{\mathcal L\left(\mathcal H^{*(a)}\right)(\tau)\right\}}{E\left(D^{(a)} \wedge \tau\right)},
$$ which includes total events, mortality, and flexible weighted combinations. Estimation uses survival-weighted Aalen-type integrals and is implemented in the `WA::LRfit()` function.

## HF-ACTION Illustrations

All three approaches were applied to the HF-ACTION dataset:

-   **Restricted WR** showed better interpretability than the original WR by standardizing follow-up to 4 years.
-   **RMT-IF** detected a significant average gain of 5.1 months in the training group, driven by both survival and reduced hospitalizations.
-   **While-alive loss** adjusted for differential follow-up and showed a 20% reduction in the average event rate, though the effect was attenuated compared to RMT-IF.

Each method offers a distinct perspective:

-   RMT-IF summarizes net benefit in time.
-   WA analyses account for differential exposure.
-   Restricted WR retains the pairwise framework but standardizes follow-up.

## Example R code

```{r}
#| eval: false
# RMT-IF
library(rmt)
obj <- rmtfit(id, time, status, trt, type = "recurrent")
summary(obj, tau = 4)

# While-alive loss (death = 2 × nonfatal)
library(WA)
obj <- LRfit(id, time, status, trt, Dweight = 2)
summary(obj, tau = 4)
```

## Conclusion

Restricting follow-up time resolves many of the interpretability issues in standard composite endpoint analysis. The restricted win ratio, restricted mean time in favor, and while-alive loss rate each target clinically interpretable, time-standardized estimands. These tools are particularly helpful when survival times vary across groups, and should be considered complementary strategies in the analysis of composite outcomes.
